### 2022-03-28 [2020-08]_ACMTOMS_(MQSI)/revisions/2022-03/response.tex

%% General paragraph settings (no indent, big line skip).
\parindent=0pt
\parskip=9pt
\baselineskip=13pt plus 1pt minus 1pt \lineskiplimit=1pt

%% Red coloring.
\input colordvi
\def\BEGINC{\textRed}
\def\ENDC{\textBlack}
\def\red#1{\BEGINC#1\ENDC}
\let\beginred=\BEGINC
\let\endred=\ENDC

%% Fonts (for title)
\font\titlebf=cmbx20
\font\titlett=cmtt20
%% Heading font.
\font\headingbf=cmbx14
\font\headingtt=cmtt14
%% Caption font.
\font\caprm=cmr8
\font\captt=cmtt8
\font\capit=cmmi8
\font\capbf=cmbx8
\font\capsy=cmsy8
%% Caption text.
\def\captext{
  \narrower
  \noindent
  \caprm
  \baselineskip=10pt plus 2pt minus 1pt
  \lineskiplimit=2pt
  \lineskip=2pt plus 1pt
  \parskip=0pt
  \textfont0=\caprm
  \textfont1=\capit
  \textfont2=\capsy
}
%% Literal "underscore" character that can be copied.
\font\underscorefont=ecrm1000
\def\_{{\underscorefont\char"5F}}

@@

{\parskip=2pt \parindent=20pt \tt
\item{} \#() \string{ echo "\# \$@" \string}
\item{} \textMaroon \# install gfortran \textBlack
\item{} sudo apt install gfortran
}

\vskip 3mm

{\tt macOS 12+} command line instructions (installs CommandLineTools and Homebrew first).

{\parskip=2pt \parindent=20pt \tt
\item{} \#() \string{ echo "\# \$@" \string}
\item{} \textMaroon \# install CommandLineTools if you do not have it already \textBlack
\item{} xcode-select -p || xcode-select --install
\item{} \textMaroon \# install Homebrew for package management \textBlack
\item{} brew --version || /bin/bash -c "\$(curl -fsSL $ $
\item{} \hskip 10mm https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
\item{} \textMaroon \# install gfortran \textBlack
\item{} brew install gfortran
}

\vskip 3mm

@@

Common instructions: Python package installation, unzipping archive, and running {\hskip 1mm} {\tt test.py}.

{\parskip=2pt \parindent=20pt \tt
\item{} \textMaroon \# install a visualization python package \textBlack
\item{} python3 -m pip install --user tlux
\item{}
\item{} \textMaroon \# unzip archive, enter resulting directory, run tests \textBlack
\item{} unzip MQSI\_testing.zip
\item{} cd MQSI\_testing
\item{} python3 test.py
\item{}
\item{} \textMaroon \# download and install R (if comparisons with Schumaker are desired) \textBlack
\item{} echo {\char13}Visit https://cran.r-project.org and follow installation instructions.{\char13}
\item{} \textMaroon \# download install Schumaker package for R \textBlack
\item{} curl --output schumaker.tar.gz $ $
\item{} \hskip 10mm https://cran.r-project.org/src/contrib/schumaker\_1.2.1.tar.gz
\item{} R CMD INSTALL schumaker.tar.gz
\item{} \textMaroon \# install a Python package that wraps calls into R \textBlack
\item{} python3 -m pip install --user rpy2
\item{}
\item{} \textMaroon \# make any desired modifications to test data at bottom of \textBlack test.py 
}

### 2022-03-30 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex

>>>>

\beginred
The binary search procedure provably converges on the boundary of the
region of monotonicity precisely defined by Ulrich and Watson [1994]
through the application of a Boolean function that is guaranteed to be
true at one end of an interval. Since a local quadratic interpolant is
applied for initial function derivative estimates, the approximation
order of the resulting fit is $O(h^3)$ (as for any second order
approximation), but it should be noted that asymptotic approximation
order is rarely of importance when considering the scattered sparse
data that $C^2$ approximations like {\tt MQSI} provide. Were exact
evaluations provided, higher order methods always win \textRed (all
other things being equal) as the data density increases. For sparse
data, what constitutes a better fit is either subjective or dependent
on the problem.
\endred

### 2022-03-30 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex

%% No extra knots, rational functions, sufficiency and necessity
%% proven, works for unequally spaced data, no code associated with
%% the paper, no algorithm provided for the initialization for free
%% shape parameters.
\refj{Abbas M., Majid, A. A. and Ali,
  J. M.}{2012}{Monotonicity-preserving $C^2$ rational cubic spline for
  monotone data}{Applied Mathematics and
  Computation}{219(6):2885--2895}

@@

% berglund2009planning
\refj{Berglund, T., Brodnik, A., Jonsson, H., Staffanson, M., and
  Soderkvist, I.}{2009}{Planning smooth and obstacle-avoiding B-spline
  paths for autonomous mining vehicles}{IEEE Transactions on
  Automation Science and Engineering}{7(1):167--172}

@@

% cameron2019moana
\refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth,
  U., Bernard, J., Jearls, C., Lux, T. C. H., Hong, Y., Watson, L. T.,
  and Butt, A. R.}{2019}{MOANA: modeling and analyzing I/O variability
  in parallel system experimental design}{IEEE Transactions on
  Parallel and Distributed Systems}{30(8):1843--1856}

%% Proves existence of splines that maintain monotonicity and
%% convexity while having fixed continuity and order.
\refj{Costantini, P.}{1986}{On monotone and convex spline
  interpolation}{Mathematics of Computation}{46(173):203--214}

%% Source paper accompanying the BVSPIS publication.
\refj{Costantini, P.}{1997}{Boundary-valued shape-preserving
  interpolating splines}{ACM Transactions on Mathematical Software
  (TOMS)}{23(2):229--251}

%% BVSPIS Algorithm.
\refj{Costantini, P.}{1997}{Algorithm 770: BVSPIS—a package for
  computing boundary-valued shape-preserving interpolating
  splines}{ACM Transactions on Mathematical Software (TOMS)}{23(2):252-–254}

%% Constructs C3 interpolating splines by progressively increasing
%% tension through Bezier control polygons until shape parameters are
%% satisfied, no code is provided.
\refj{Cravero, I., C. Manni}{2003}{Shape-preserving interpolants with
  high smoothness}{Journal of computational and applied
  mathematics}{157(2):383--405}

%% Iterates towards a solution of nonlinear equations that ensure C2
%% continuity of quadratic rational splines. The solution they iterate
%% towards exists and is unique. They provide proofs and methods for
%% setting derivatives at the ends, no formal algorithms or code.
\refj{Delbourgo, R., Gregory, J. A.}{1983}{$C^2$ rational quadratic
  spline interpolation to monotonic data}{IMA Journal of Numerical
  Analysis}{3(2):141--152}

%% Continuous previous work by incorporating a cubic numerator, proves
%% shape preservation when the correct value for a tension parameter
%% is set but no method for setting that parameter. Accuracy depends
%% on accurate derivative evaluations. No code nor formal algorithm is
%% provided.
\refj{Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in
  tension}{SIAM Journal on Numerical Analysis}{30(2):595--607}

%% Proves approximation error bounds for monotone spline interpolation
%% of data from monotone functions.
\refj{DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM
  Journal on Mathematical Analysis}{8(5):891--905}

%% Accompanies the previous error bounds in the context of monotone
%% polynomial interpolation of data from a monotone function.
\refj{DeVore, R. A.}{1977}{Monotone approximation by polynomials}{SIAM
  Journal on Mathematical Analysis}{8(5):906--921}

%% Proposes an algorithm for constructing C2 Hermite monotone spline
%% interpolants using proven sufficient, but not necessary
%% conditions. As these bounds are not sharp, the space of allowable
%% derivative and second derivative values at knots is unnecessarily
%% shrunk. An algorithm is provided, but no code. Recommends against
%% the direct optimization for smoothness parameters like L2, instead
%% promoting problem specific definitions of 'geometric niceness'.
\refj{Dougherty, R. L., Edelman, A., Hyman,
  J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving
  cubic and quintic Hermite interpolation}{Mathematics of
  Computation}{52(186):471--494}

%% Proves a simple method for determining the existence of a monotone
%% C2 cubic interpolating spline, but functions of this kind do not
%% always exist for arbitrary convex monotone data sets.
\refj{Fiorot, J. C., Tabka, J.}{1991}{Shape-preserving $C^2$ cubic
  polynomial interpolating splines}{Mathematics of
  Computation}{57(195):291--298}

@@

%% Proves necessary and sufficient conditions for a closed form
%% solution to monotone C1 quadratic rational spline
%% interpolation. Does not provide formal algorithm, nor code.
\refj{Gregory , J. A.,Delbourgo, R.}{1982}{Piecewise rational
  quadratic interpolation to monotonic data}{IMA Journal of Numerical
  Analysis}{2(2):123--130}

% gregory1985shape
\reft{Gregory, J. A.}{1985}{Shape preserving spline
  interpolation}{19850020252, Brunel University}

%% Proposes a method for making arbitrarily smooth shape preserving
%% spline interpolants, but uses a centered quadratic to estimate
%% derivatives and a sufficient set of bounding conditions to ensure
%% shape preservation in practice. No necessary conditions are
%% applied, no algorithm is provided, neither is any code provided.
\refj{Han, X.}{2018}{Shape-preserving piecewise rational interpolation
  with higher order continuity}{Applied Mathematics and
  Computation}{337:1--13}

% herman2015techniques
\reft{Herman, D. L., and Oftedal, M. J.}{2015}{Techniques and
  workflows for computer graphics animation systems}{US Patent
  9,216,351}

@@

%% Presents several nonlinear boundary conditions based on the median
%% function to establish a fourth order monotonicity preserving
%% constraints relying on divided difference initial derivative
%% estimates. Overall generates a monotone C1 cubic interpolant to
%% data.
\refj{Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM
  Journal of Numerical Analysis}{30(1):57--100}

@@

% lux2020thesis
\reft{Lux, T. C. H.}{2020}{Interpolants and error bounds for modeling
  and predicting variability in computer systems}{Thesis, Virginia
  Polytechnic Institute and State University, Blacksburg, Virginia}

%% Constructs a C2 cubic spline with two additional knots per
%% interval. No code (paywall).
\refj{Manni, C., Sablonni\`ere, P.}{1997}{Monotone interpolation of
  order 3 by $C^2$ cubic splines}{IMA Journal of Numerical
  Analysis}{17(2):305--320}

%% Uses sufficient conditions to generate C2 cubic spline. No code
%% (paywall).
\refj{Manni, C.}{2001}{On shape preserving $C^2$ Hermite
  interpolation}{BIT Numerical Mathematics}{41(1):127--148}

% mcallister1981algorithm
\refj{McAllister, D. F. and Roulier, J. A.}{1981}{An algorithm for 
  computing a shape-preserving osculatory quadratic spline}{ACM 
  Transactions on Mathematical Software (TOMS)}{7(3):331--347}

% mcallister1981shape
\refj{McAllister, D. F. and Roulier, J. A.}{1981}{Algorithm 574:
  Shape-Preserving Osculatory Quadratic Splines [E1, E2]}{ACM
  Transactions on Mathematical Software (TOMS)}{7(3):384--386}

@@

@@

%% Proposes a cubic C2 shape preserving spline method with free
%% parameters (mentions quintic splines are necessary for globally C2
%% monotone splines). Sufficient (but not necessary) conditions are
%% provided for monotonicity.
\refj{Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline
  interpolation}{IMA Journal of Numerical Analysis}{13(4):493--507}

@@

@@

% yao2018unconditionally
\refj{Yao, J. and Nelson, K. E.}{2018}{An unconditionally monotone
  $\scriptstyle C^2$ quartic spline method with nonoscillation
  derivatives}{Advances in Pure Mathematics}{8(LLNL-JRNL-742107)}

%% Uses sufficient conditions over three free parameters to construct
%% C2 rational quartic shape preserving splines.
\refj{Zhu, Y.,Han, X.}{2015}{$C^2$ rational quartic interpolation
  spline with local shape preserving property}{Applied Mathematics
  Letters}{46:57--63}

%% Uses sufficient conditions over two free parameters to construct C2
%% rational quartic shape preserving splines.
\refj{Zhu, Y.,Han, X.}{2015}{Shape preserving $C^2$ rational quartic
  interpolation spline with two parameters}{International Journal of
  Computer Mathematics}{92(10):2160--2177}

}
%% -------------------------------------------------------------------

>>>>

%% No extra knots, rational functions, sufficiency and necessity
%% proven, works for unequally spaced data, no code associated with
%% the paper, no algorithm provided for the initialization for free
%% shape parameters.
\refj{\beginred Abbas M., Majid, A. A. and Ali,
  J. M.}{2012}{Monotonicity-preserving $C^2$ rational cubic spline for
  monotone data}{Applied Mathematics and
  Computation}{219(6):2885--2895}
\endred

@@

% berglund2009planning
\refj{Berglund, T., Brodnik, A., Jonsson, H., Staffanson, M. and
  Soderkvist, I.}{2009}{Planning smooth and obstacle-avoiding B-spline
  paths for autonomous mining vehicles}{IEEE Transactions on
  Automation Science and Engineering}{7(1):167--172}

@@

% cameron2019moana
\refj{Cameron, K. W., Anwar, A., Cheng, Y., Xu, L., Li, B., Ananth,
  U., Bernard, J., Jearls, C., Lux, T. C. H., Hong, Y., Watson, L. T.
  and Butt, A. R.}{2019}{MOANA: modeling and analyzing I/O variability
  in parallel system experimental design}{IEEE Transactions on
  Parallel and Distributed Systems}{30(8):1843--1856}

%% Proves existence of splines that maintain monotonicity and
%% convexity while having fixed continuity and order.
\refj{\beginred Costantini, P.}{1986}{On monotone and convex spline
  interpolation}{Mathematics of Computation}{46(173):203--214}
\endred

%% Source paper accompanying the BVSPIS publication.
\refj{\beginred Costantini, P.}{1997a}{Boundary-valued shape-preserving
  interpolating splines}{ACM Transactions on Mathematical Software
  (TOMS)}{23(2):229--251}
\endred

%% BVSPIS Algorithm.
\refj{\beginred Costantini, P.}{1997b}{Algorithm 770: BVSPIS -- A package for
  computing boundary-valued shape-preserving interpolating
  splines}{ACM Transactions on Mathematical Software
  (TOMS)}{23(2):252-–254}
\endred

%% Constructs C3 interpolating splines by progressively increasing
%% tension through Bezier control polygons until shape parameters are
%% satisfied, no code is provided.
\refj{\beginred Cravero, I. and C. Manni}{2003}{Shape-preserving interpolants with
  high smoothness}{Journal of computational and applied
  mathematics}{157(2):383--405}
\endred

%% Iterates towards a solution of nonlinear equations that ensure C2
%% continuity of quadratic rational splines. The solution they iterate
%% towards exists and is unique. They provide proofs and methods for
%% setting derivatives at the ends, no formal algorithms or code.
\refj{\beginred Delbourgo, R. and Gregory, J. A.}{1983}{$C^2$ rational quadratic
  spline interpolation to monotonic data}{IMA Journal of Numerical
  Analysis}{3(2):141--152}
\endred

%% Continuous previous work by incorporating a cubic numerator, proves
%% shape preservation when the correct value for a tension parameter
%% is set but no method for setting that parameter. Accuracy depends
%% on accurate derivative evaluations. No code nor formal algorithm is
%% provided.
\refj{\beginred Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in
  tension}{SIAM Journal on Numerical Analysis}{30(2):595--607}
\endred

%% Proves approximation error bounds for monotone spline interpolation
%% of data from monotone functions.
\refj{\beginred DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM
  Journal on Mathematical Analysis}{8(5):891--905}
\endred

%% Proposes an algorithm for constructing C2 Hermite monotone spline
%% interpolants using proven sufficient, but not necessary
%% conditions. As these bounds are not sharp, the space of allowable
%% derivative and second derivative values at knots is unnecessarily
%% shrunk. An algorithm is provided, but no code. Recommends against
%% the direct optimization for smoothness parameters like L2, instead
%% promoting problem specific definitions of 'geometric niceness'.
\refj{\beginred Dougherty, R. L., Edelman, A. and Hyman,
  J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving
  cubic and quintic Hermite interpolation}{Mathematics of
  Computation}{52(186):471--494}
\endred

%% Proves a simple method for determining the existence of a monotone
%% C2 cubic interpolating spline, but functions of this kind do not
%% always exist for arbitrary convex monotone data sets.
\refj{\beginred Fiorot, J. C. and Tabka, J.}{1991}{Shape-preserving $C^2$ cubic
  polynomial interpolating splines}{Mathematics of
  Computation}{57(195):291--298} \endred \goodbreak

@@

%% Proves necessary and sufficient conditions for a closed form
%% solution to monotone C1 quadratic rational spline
%% interpolation. Does not provide formal algorithm, nor code.
\refj{\beginred Gregory, J. A. and Delbourgo, R.}{1982}{Piecewise rational
  quadratic interpolation to monotonic data}{IMA Journal of Numerical
  Analysis}{2(2):123--130}
\endred

% gregory1985shape
\reft{Gregory, J. A.}{1985}{Shape preserving spline
  interpolation}{NASA, Langley Research Center Computational Geometry
  and Computer-Aided Design}

% The paper where a linear facet model is proposed.
\refj{\beginred Haralick, R. M. and Watson, L. T.}{1981}{A facet model for image
  data}{Computer Graphics Image Processing}{15:113--129}
\endred

%% Proposes a method for making arbitrarily smooth shape preserving
%% spline interpolants, but uses a centered quadratic to estimate
%% derivatives and a sufficient set of bounding conditions to ensure
%% shape preservation in practice. No necessary conditions are
%% applied, no algorithm is provided, neither is any code provided.
\refj{\beginred Han, X.}{2018}{Shape-preserving piecewise rational interpolation
  with higher order continuity}{Applied Mathematics and
  Computation}{337:1--13}
\endred

% herman2015techniques
\reft{Herman, D. L. and Oftedal, M. J.}{2015}{Techniques and
  workflows for computer graphics animation systems}{US Patent
  9,216,351}

@@

%% Presents several nonlinear boundary conditions based on the median
%% function to establish a fourth order monotonicity preserving
%% constraints relying on divided difference initial derivative
%% estimates. Overall generates a monotone C1 cubic interpolant to
%% data.
\refj{\beginred Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM
  Journal of Numerical Analysis}{30(1):57--100}
\endred

@@

% lux2020thesis
\reft{Lux, T. C. H.}{2020}{Interpolants and error bounds for modeling
  and predicting variability in computer systems}{Ph.D. thesis,
  Department of Computer Science, Virginia Polytechnic Institute and
  State University, Blacksburg, Virginia}

% Monotone quintic algorithm.
\refj{\beginred Lux, T. C. H., Watson, L. T., Chang, T. H., Xu, L.,
  Wang, Y. and Hong, Y.}{2020}{An algorithm for constructing monotone
  quintic interpolating splines}{SCS {\sl Spring Simulation
    Multiconference, High Performance Computing Symposium}}{Article
  33, 1--12}
\endred

%% Constructs a C2 cubic spline with two additional knots per
%% interval. No code (paywall).
\refj{\beginred Manni, C. and Sablonni\`ere, P.}{1997}{Monotone interpolation of
  order 3 by $C^2$ cubic splines}{IMA Journal of Numerical
  Analysis}{17(2):305--320}
\endred

%% Uses sufficient conditions to generate C2 cubic spline. No code
%% (paywall).
\refj{\beginred Manni, C.}{2001}{On shape preserving $C^2$ Hermite
  interpolation}{BIT Numerical Mathematics}{41(1):127--148}
\endred

% mcallister1981algorithm
\refj{McAllister, D. F. and Roulier, J. A.}{1981a}{An algorithm for 
  computing a shape-preserving osculatory quadratic spline}{ACM 
  Transactions on Mathematical Software (TOMS)}{7(3):331--347}

% mcallister1981shape
\refj{McAllister, D. F. and Roulier, J. A.}{1981b}{Algorithm 574:
  Shape-Preserving Osculatory Quadratic Splines [E1, E2]}{ACM
  Transactions on Mathematical Software (TOMS)}{7(3):384--386}

@@

% Proves that C2 convexity preserving splines do not always exist.
\reft{\beginred Mulansky, B. and Neamtu, M.}{1991}{On the existence of shape
  preserving interpolation operators}{Dissertation, University of
  Twente, Vancouver}
\endred

@@

%% Proposes a cubic C2 shape preserving spline method with free
%% parameters (mentions quintic splines are necessary for globally C2
%% monotone splines). Sufficient (but not necessary) conditions are
%% provided for monotonicity.
\refj{\beginred Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline
  interpolation}{IMA Journal of Numerical Analysis}{13(4):493--507}
\endred

@@

% ulrich1990positivity tech report
\reft{\beginred Ulrich, G. and Watson, L. T.}{1990}{Positivity conditions for
  quartic polynomials}{Technical Report 90-57, Department of Computer
  Science, Virginia Polytechnic Institute and State University,
  Blacksburg, Virginia}
\endred

@@

% yao2018unconditionally
\refj{Yao, J. and Nelson, K. E.}{2018}{An unconditionally monotone
  $\scriptstyle C^2$ quartic spline method with nonoscillation
  derivatives}{Advances in Pure Mathematics}{8(1):25--40}

%% Uses sufficient conditions over three free parameters to construct
%% C2 rational quartic shape preserving splines.
\refj{\beginred Zhu, Y. and Han, X.}{2015a}{$C^2$ rational quartic interpolation
  spline with local shape preserving property}{Applied Mathematics
  Letters}{46:57--63}
\endred

%% Uses sufficient conditions over two free parameters to construct C2
%% rational quartic shape preserving splines.
\refj{\beginred Zhu, Y. and Han, X.}{2015b}{Shape preserving $C^2$ rational quartic
  interpolation spline with two parameters}{International Journal of
  Computer Mathematics}{92(10):2160--2177}
\endred
}
%% -------------------------------------------------------------------
